Title Pending

### 2019-02-18 manuscript/manuscript.tex
reorganization of section 2

@@

\subsection{Pessimistic Inference}

In this first section, we demonstrate how learning under an assumption of a lack of future control naturally leads to pessimistic inference and second-order conditioning. A symptom central to all of the anxiety disorders is pessimistic inference, or a tendency to appraise the likelihood and/or severity of threatening situations out of proportion to their actual environmental contingencies \citep{dsm5, BeckClark1997, ClarkBeck2011}. For example, an individual with specific phobia may vastly exaggerate the probability of encountering their fear (e.g. a venomous spider) in their everyday lives. Anxious individuals will also tend to associate fear with cues tangentially related to signals of threat. For example, the same individual may begin to fear locations where they may encounter spider webs (e.g. basements).

Pessimistic inference in anxiety is well-documented. In studies of self-reported beliefs, individuals with anxiety rate hypothetical future negative outcomes as being more likely to occur and worse to experience than non-anxious counterparts \citep{ButlerMathews1983, ButlerMathews1987, Foa1996, MacLeod1996, MacLeod1997, Luten1997, Stober1997, Borkovec1999, Maner2006, Mitte2007, Miranda2007}. In the laboratory, exaggerated threat appraisal in anxiety has traditionally been measured through fear conditioning. The results of these studies demonstrate that anxious individuals exhibit greater fear towards threatening cues, non-threatening cues, and extinguished threatening cues as compared to non-anxious counterparts \citep{lissek2005, MinekaOehlberg2008, Duits2015}. Importantly, threatening cues can become associated with secondary neutral cues, allowing negative value to spread \citep{wessa2007}. In sum, anxiety is associated with inflated estimates of threat, both likelihood and value, and these estimates can spread from the source to distantly associated cues.

The same cannot be said for a pessimistic agent. If an agent doubts its future ability to act in according with the optimal policy, then it cannot rule out that, in the future, it may inadvertently encounter the harmful state. This is more so likely if the agent is already proximal to the harmful state. Under this pessimistic belief, the possibility of future harm is no longer zero, as it is for the optimistic agent, and thus negative value can propagate from the states antecedent to harm, back to their respective antecedents and so on. The extent of this spread will depend on the strength of the pessimistic belief (Figures 1c, 1d). With increasing pessimism, harmful outcomes are assigned greater credence and thus exert greater and wider influence on antecedent states.

In sequential decision environments then, a perceived lack of self-efficacy can result in the propagation of negative value from potential future harms. Such a pessimistic agent may correspondingly report an increased likelihood of threat (i.e. a greater number of threatening states), increased severity of threat (i.e. states with more negative value), and fear of states antecedent to the fear source. As such, pessimistic learning can account for two of the hallmark symptoms of the anxiety disorders.

\subsection{Avoidance}

Avoidance behaviors in anxiety have been studied using a variety of methods. Perhaps the most classic paradigm is the behavioral avoidance test \citep{bandura1977}, wherein participants complete a number of tasks involving a feared object or situation and the number of tasks completed (or lack thereof) is a measure of avoidance. Computerized avoidance tasks have become increasingly common, and increased distance from virtual threat has been found to correlate with anxiety \citep{Bach2014, Bach2017, Sheynin2014}.

There is a long history of models attempting to explain avoidance learning. The fundamental challenge is explaining how an action can be reinforced in the absence of an outcome. Two-factor theory was the explanation for a long while, but suffered from a number of noteworthy criticisms \citep{Krypotos2015}. Recent work has revitalized two-factor theory and integrated it with reinforcement learning \citep{Moutoussis2008, Maia2010}, and in the process addressing several criticisms of two-factor theory. As has been discussed elsewhere \citep{Moutoussis2017}, such accounts only explain the maintenance of avoidance in short timescales, but do not explain why resistant to change and why they do not go away in general.

\begin{figure}
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../figures/02_cliff.png}}%
  }
  \par \textbf{Figure 2:} Cliff-walking is a deterministic gridworld where an agent must navigate from one side (S) to the goal on the other side (G) (A). Every step costs the agent ($r=-1$) but stepping off the cliff results in great harm ($r=-100$). An optimistic agent learns to travel the shortest (loss-minimizing) path along the cliff (B). For a pessimistic agent ($w=0.8$), the potential for harm yields an optimal policy that maintains distance from the cliff's edge (C). With increasing pessimism ($w=0.6$), the agent learns an optimal path off the cliff so as to minimize unnecessary steps prior to its perceived inevitable harm (D). (Parameters: $\gamma = 1$)
\end{figure}

In our second simulation, we demonstrate how pessimistic learning can yield sustained avoidance. In the cliff-walking environment \citep{SuttonBarto1998, SuttonBarto2018, Gaskett2003}, an agent must navigate from a start on side of the map to the other (Figure 2a). The agent experiences a small penalty for every step it takes. The shortest (least costly) path is along the cliff's edge; if the agent steps off the cliff, however, it incurs a much larger penalty. As in the first environment, we assume a deterministic transition structure.

Just as in the first environment, the long-run value estimates by an optimistic agent do not reflect the potential harm of the cliff (Figure 2b). Because this agent (rightfully) expects to act according to the optimal policy in the future (i.e. never step off the cliff), the agent perceives no risk of future disaster and thus the large penalties of stepping off the cliff do not pollute the agent's value estimates. The optimistic agent learns the loss-minimizing policy and walks along the cliff's edge.

This is not the case for the pessimistic agent. As above, when the agent doubts its future ability to act in a reward-maximizing (loss-minimizing) manor, it cannot rule out the possibility of accidentally stepping off the cliff. Thus, the potential for future harm is no longer zero and the large penalties for stepping off the cliff may back-propagate to their antecedent states, and so on. The extent of this spread is dependent on the strength of the pessimistic belief. With increasing pessimism, the optimal policy for the agent is increasingly distance from the cliff's edge (Figure 2c). Surprisingly, at high levels of pessimistic belief, the optimal policy is to immediately step off the cliff (Figure 2d). The reason for this is simple: with high levels of pessimism, the likelihood of disaster grows large enough that any step is simply viewed as adding onto the pain.

Under pessimistic learning then, a perceived lack of self-efficacy can yield the propagation of negative value from potential future harms. In such a scenario, the optimal policy for an agent is to maintain a distance from threat far enough so as to optimally balance the risk of catastrophe with foregone losses. Importantly, these simulations demonstrate the long-run value estimates rather than any momentary value estimates. In other words, the behavioral policy these agents demonstrate will be maintained insofar beliefs about self-efficacy are maintained, irrespective of any momentary noisiness in actual outcomes. Thus, pessimistic learning can account for pathological avoidance in anxiety disorders.

>>>>

The consequences of this manipulation for the open field environment are presented in Figure 1c,d. When Bellman optimality is violated (i.e. an agent no longer assumes the value of a successor state is equal the best available action), more than reward opportunities are able to propagate value through the state space. Indeed when $w$ is non-zero, the agent treats the future potential of danger as possible and thus negative value is able to spread from its source recursively through antecedent states thereby degrading the value of distal states (the degree of which determined by the extent of pessimism, $w$, and temporal discounting, $\gamma$). With increasing certainty that the worst policy will be followed in the future, the larger the extent to which threat value can permeate.

In addition to drawing inspiration from clinical theories of anxiety, this pessimistic model also captures several core symptoms of anxiety disorders. The hallmark symptom of anxiety disorders is exaggerated threat appraisal, or the tendency to perceive threat as disproportionately greater in likelihood and severity than is justified. This is evident in the anxious state value estimates presented in Figure 1c,d. Indeed when safety is perceived to be uncertain, threat in the environment is exaggerated in magnitude and prevalence. As should be apparent, this very same mechanism captures the generalization of fear that often accompanies anxiety disorders.

Avoidance behaviors in anxiety have been studied using a variety of methods. Perhaps the most classic paradigm is the behavioral avoidance test \citep{bandura1977}, wherein participants complete a number of tasks involving a feared object or situation and the number of tasks completed (or lack thereof) is a measure of avoidance. Computerized avoidance tasks have become increasingly common, and increased distance from virtual threat has been found to correlate with anxiety \citep{Bach2014, Bach2017, Sheynin2014}. This is not the case for the pessimistic agent. As above, when the agent doubts its future ability to act in a reward-maximizing (loss-minimizing) manor, it cannot rule out the possibility of accidentally stepping off the cliff. Thus, the potential for future harm is no longer zero and the large penalties for stepping off the cliff may back-propagate to their antecedent states, and so on. The extent of this spread is dependent on the strength of the pessimistic belief. With increasing pessimism, the optimal policy for the agent is increasingly distance from the cliff's edge (Figure 2c). Surprisingly, at high levels of pessimistic belief, the optimal policy is to immediately step off the cliff (Figure 2d). The reason for this is simple: with high levels of pessimism, the likelihood of disaster grows large enough that any step is simply viewed as adding onto the pain.

@@
